---
title: "ml_a1_md_draft"
author: "Josh Grosman and Tim Kirkbride"
date: "March 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
library(mlr)
library(lubridate)
```

## Introduction

This project aimed to construct a classification model to predict whether a Kickstarter project will be successful or otherwise. This project is split into two phases: Phase 1, which will focus on data preprocessing and exploration, and Phase 2, which will be concered with model building and validation. Only Phase 1 of the project will be presented in this report. In this phase, the dataset and its attrbiutes will be unpacked and described. The data will then be cleaned and preprocessed and finally, all relevant attrbiutes will be explored with regards to their relationships with each other.

## Dataset

```{r}
ks <- read_csv("ks-projects-201801.csv")
```

The dataset used was obtained from Kaggle and can be found here:

https://www.kaggle.com/kemical/kickstarter-projects/data

This project used the 2018 dataset, which contained the more recently updated data. The dataset contained 378,661 observations, each corresponding to a different Kickstarter project launched between 2009-2018. The dataset consisted of 14 feature variables and one target variable. 


### Target Feautre

The target feature used in this analysis was the state of the project at the time of its deadline. The levels of the target variable are denoted as follows:

* Cancelled: If the project was cancelled premature to its deadline

* Failed: If the project was unable to generate its goal amount by the time of its deadline

* Live: If the project has not yet reached its deadline and is still active

* Succesful: If the project was able to at least generate its goal amount by the time of its deadline

* Suspended: If the project is found to be in breech of the Kickstarter rules, all funding for the project stops indefintely and all previous pledges are cancelled

* Undefined: If it is unknown which of the above categories is most appropriate

While the target feature was originally comprised of six classes, the present analysis was only concerned with predicting whether a Kickstarter project will be successful or fail. Furthermore, all of the observations which related to the four remaining target levels combined only comprised a minority of the overall observations (10.91%). As such, it was decided that observations classifed under the levels of live, canceled, suspended or undefined would not be considered in the analysis, and only observations relating to successful or failed projects would be retained.

```{r}
ks_filt<-ks%>%filter(state %in% c("successful","failed"))
```

As as result of this, the dataset now contained 331,675 observations relating to successful and failed Kickstarter projects. The target feature would only have two classes and therefore the analysis would be binary classifcation.

### Descriptive Features

* ID: continuous
* name: string
* category: factor (159 levels)
* main_category: factor (Art, Comics, Crafts, Dance, Design, Fashion, Film & Video, Food, Games, Journalism, Music, Photography, Publishing, Technology, Theater)
* currency: factor (AUD, CAD, CHF, DKK, EUR, GBP, HKD, JPY, MXN, NOK, NZD, SEK, SGD, USD)
* deadline: date
* goal: continuous
* launched: date
* pledged: continuous
* backers: integer
* country: (AT, AU, BE, CA, CH, DE, DK, ES, FR, GB, HK, IE, IT, JP, LU, MX, N,0\", NL, NO, NZ, SE, SG, US)
* us_pledged: continuous
* us_pledged_real: continuous
* us_goal_real: continuous

Many of the descriptive features here are self-explanatory. The last three features are corrections of the pledged and goal features, which convert the relevant currency amounts of the associated observations to USD. Information from the website where the dataset was retrived suggests that usd_pledged was created to standardise the pledged column to USD, but there were issues with the conversion which resulted in errors. The usd_pledged_real feature amended these errors and has the correct amount pledged in USD. The feature usd_goal_real has the USD amount of the project's goal.

Inspection of the levels of the country feature highlight a potential typo, where a country was assigned the code 'N,0\"', which clearly does not correspond with the rest of the levels. This will be examined further in the next section of this report.

## Data Preprocessing

### Initial Inspection and Error Identification

Prior to summarization, some obviously redundant columns were removed. The features ID and name would clearly add no value in predicting the target variable and were dropped. Additionally, as the columns usd_goal_real and usd_pledged_real contained the correct standardised goal and pledged amounts respectively, the pledged, goal and usd_pledged features were no longer relevant and were dropped.

```{r}
ks_filt$ID<-NULL
ks_filt$name<-NULL
ks_filt$`usd pledged`<-NULL
ks_filt$pledged<-NULL
ks_filt$goal<-NULL
```

The dataset was summarised in Table 1 and 2 below.

```{r}
head(ks_filt)%>% knitr::kable( caption = 'Table 1. First 6 rows of Kickstarter Data')
```


```{r}
summarizeColumns(ks_filt)%>% knitr::kable( caption = 'Table 2. Feature Summary before Data Preprocessing')

```

From Table 1 and Table 2, it is clear that data has been read in correctly and there no missing values or any obvious errors/typos. However, some issues were noted.

* As mentioned previously, there was a suspected typo in the name of one the levels of the country feature - N,0\".

* Looking at the classes of each feature shown in Table 1, the two date columns - launched and deadline - are not of the same class, which may become an issue if these columns are needed to created a new variable. Furthermore, as the time dimension is applicable as a feature within the present analysis, these two columns will ultimately need to be dropped.

* The catergory and main_category columns appear to represent the same underlying feature - the type of buisness of the Kickstarter project, however, as seen in Table 2, the cardinality of category was very large at 159 levels compared to that of main_category, which had only 15 levels.

* The currency and country columns seem to hold fairly similar information. As such, dropping one may be necessary. This would need to be investigated visually, however, so will be dealt with in the next section.

* It is clear that all factor columns were defined as the character class when imported and will need be converted.

### Data Wrangling and Correction

Firstly, the extent of the N,0\" typo within the country feature was explored further. Perusal of the website where data were retreived provided no explaination to the meaning of this value. Therefore, this level was considered to classify as a missing value, and was converted as such. The proportion of complete cases (i.e. those which had no missing values) was then calculated.

```{r}
ks_filt$country <- ifelse(ks_filt$country == "N,0\"", NA, ks_filt$country)
sum(complete.cases(ks_filt))/nrow(ks_filt)
```

Here, we see that the proprotion of complete cases is very high, at 99.94% of all observations. This suggested that the number of observations affected by the typo was minimal and as such, it was deemed appropriate to omit all affected observations from the analysis.

```{r}
ks_filt<-ks_filt%>%na.omit()
```

After this, the number of observations became 331,465.

Next, to was decided that a new feature would be created to represent the duration in days that each project lasted by finding the difference between the two date columns. First, the launched column would have to be converted to the correct date format. In its current state, launched included not only the date, but also the time at which the project was launched. As such, it was converted.

```{r}
ks_filt$launched <- format(as.POSIXct(ks_filt$launched,format='%Y/%m/%d %H:%M:%S'),format='%Y/%m/%d')
ks_filt$launched<-ks_filt$launched%>%as.Date()
```

The duration feature was then added to the dataset by subtracting the launched column from the deadline column. Subseqently, the launched and deadline columns were dropped as they were no longer needed.

```{r}
ks_filt$duration<-as.numeric(ks_filt$deadline-ks_filt$launched)
ks_filt$launched<-NULL
ks_filt$deadline<-NULL
```

Next, the main category and category features were considered. Inspection of these features highlighted that main_category was simply be a reduced version of catgeory with far fewer levels. This left category essentially redundant, so it was removed.

```{r}
ks_filt$category<-NULL
```

Finally, all features loaded in as the character type were converted to factors.

```{r}
ks_filt[sapply(ks_filt, is.character)] <- lapply(ks_filt[sapply(ks_filt, is.character)], as.factor)
```



# Data Exploration

## Univariate Visualisations

Initially, the overall counts of succesful and failed projects was examined.

```{r}
binary<-ggplot(ks_filt,aes(x=state))

binary+geom_bar()+
  labs(title="Bar Chart of Project States")
```

Clearly the main point here is that there were more failed projects in the data than there were sucessful ones. This should be considered in visualisations to come.

The main_category feature was then visualised to explore the differences across its levels with regards to the state of the project.

```{r}
cat <- ggplot(ks_filt, aes(x = main_category))

cat + geom_bar()+
  labs(title = "Bar Chart of Project Category",
       x = "Category",
       y = "Count")+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```


```{r}
cat_state <- table(ks_filt$main_category,ks_filt$state, dnn = c("Category","State"))
prop.cat_state<-data.frame(prop.table(cat_state,1))
colnames(prop.cat_state) <- c("Category","State","Proportion")

p5<-ggplot(prop.cat_state,aes(x=Category,y=Proportion,fill=State))
p5+geom_bar(stat="identity",position="dodge")+
  labs(title="Proportions of Successful/Failed Projects within Project Catgeories",
       y="Proportion within Category")+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

Here we see the distribution of the types of projects included in the dataset, as well as the proportions of successful and failed projects within each category. Showing proportions instead of counts allows for direct comparison accross the levels of the main_category feature. This highlights that generally, most categories are associted with failed projects, except for comics, dance, music and theater. Clearly these are all related to the arts.

Next we looked at the currency and country feature simultaniously to determine if either was redundant.

```{r}
ggplot(ks_filt, aes(x = country)) + geom_bar() +
  labs(title="Bar Chart of Country",
       y="Count",
       x="Country")
```

```{r}
ggplot(ks_filt, aes(x = currency)) + geom_bar()+
  labs(title="Bar Chart of Currency",
       y="Count",
       x="Country")
```

Inspection of these two plots does highlight a striking similarity. Additionally, they also serve to emphasise that a large proportion of the projects are associated with the US, and also, that many other countries and currencies have much smaller amounts of associated projects. As such, utilising either of these features may lead to some issues. It was therefore decided that it would be best to aggregate the levels of the country feature to represent the continents, and to then drop the country and currency features. The plots of the new continent feature are shown below.

```{r}
europe <- c("AT", "BE", "DE", "DK", "ES", "FR", "GB", "IE", "IT", "LU", "NL", "NO", "SE")
north_america <- c("CA", "MX", "US")
asia <- c("CH", "HK", "JP", "SG")
oceania <- c("AU", "NZ")

ks_filt$continent <- case_when(
  ks_filt$country %in% europe ~ "Europe",
  ks_filt$country %in% north_america ~ "North America",
  ks_filt$country %in% asia ~ "Asia",
  ks_filt$country %in% oceania ~ "Oceania"
)

ks_filt$country<-NULL
ks_filt$currency<-NULL
```
```{r}
cont <- ggplot(ks_filt, aes(x = continent))

cont + geom_bar(position = "dodge")+
  labs(title="Bar Chart of Continent",
       y="Count",
       x="Continent")
```

Obviously, this chart still shows that North America dominates the projects being assessed, however the other levels are now more concentrated. Despite this, the amount of projects originating from Asia and Oceania remains very small.

```{r}
cont_state <- table(ks_filt$continent,ks_filt$state, dnn = c("Continent","State"))
prop.cont_state<-data.frame(prop.table(cont_state,1))
colnames(prop.cont_state) <- c("Continent","State","Proportion")

p6<-ggplot(prop.cont_state,aes(x=Continent,y=Proportion,fill=State))
p6+geom_bar(stat="identity",position="dodge")+
  labs(title="Proportions of Successful/Failed Projects within Project Continents")
```

From the chart above, we see that the proportions of failed and succesfull projects from Asia and Europe are almost identical. North America appears to have a slightly higher proportion of successful projects in comparision, while Oceania has a lower proportion of successful projects.

```{r}
g <- ggplot(ks_filt, aes(x = duration))

#g + geom_histogram()

g + geom_histogram(aes(fill = state)) + 
labs(title = "Duration of Project, by Outcome",
x = "Duration",
y = "Count",
fill = "Outcome") +
facet_grid(~state)
```

Next is backers

```{r}
# couldnt get this to work, will need to come back to it
back<-ggplot(ks_filt,aes(x=backers))

back+geom_histogram()

```


Here we can see that there is an obviouos outlier, to improve the data, we should inspect this row

```{r}
# couldnt get this to work, will need to come back to it

max(ks_filt$backers)

ks_filt %>% filter(backers == max(ks_filt$backers))

```

With further research we can see that this is a valid row (The project was the most backed of all time), for visualisation purposes however, will we remove it from the plot

```{r}
# thought overlaying density might make it clearer (also accounts for differnt proportions)
back_filt<-ks_filt%>%filter(backers<600)

backers <- ggplot(back_filt, aes(x = backers,fill=state))

backers + geom_density(alpha=.4) 

```

usd_goal_real

```{r}
goal <- ggplot(ks_filt, aes(x = usd_goal_real))

goal + geom_histogram()
```

Filtering the data 

```{r}
ks_goal <- ks_filt %>% filter(usd_goal_real <= 10000)

filt_goal <- ggplot(ks_goal, aes(x = usd_goal_real))

filt_goal + geom_histogram()

filt_goal + geom_histogram(aes(fill = state)) +
facet_grid(~state)

filt_goal + geom_density(aes(fill = state)) +
  facet_grid(~state)

filt_goal + geom_density(aes(fill = state), alpha = 0.4)
```