---
title: "ml_a1_md_draft"
author: "Josh Grosman and Tim Kirkbride"
date: "March 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
library(mlr)
library(lubridate)
```

## Introduction

This project aimed to construct a classification model to predict whether a Kickstarter project will be successful or fail. The project is split into two phases. Phase 1 will focus on data preprocessing and exploration and will be covered in this report, while Phase 2 will be concered with model building and validation and will be covered in a later report. In this phase, the dataset and its attrbiutes will be unpacked and described. The data will then be cleaned and preprocessed. Finally, all relevant attrbiutes will be explored with regards to their relationships with each other.

## Dataset

```{r}
ks <- read_csv("ks-projects-201801.csv")
```

The dataset used was obtained from Kaggle and can be found here:

https://www.kaggle.com/kemical/kickstarter-projects/data

This project used the 2018 dataset, which contained the more recently updated data. The dataset contained 378,661 observations, each corresponding to a different Kickstarter project launched between 2009-2018. The dataset consisted of 14 feature variables and one target variable. 


### Target Feautre

The target feature used in this analysis was the state of the project at the time of its deadline. The levels of the target variable are denoted as follows:

* Canceled: If the project was cancelled premature to its deadline

* Failed: If the project was unable to generate its goal amount by the time of its deadline

* Live: If the project has not yet reached its deadline and is still active

* Successful: If the project was able to at least generate its goal amount by the time of its deadline

* Suspended: If the project is found to be in breech of the Kickstarter rules, all funding for the project stops indefintely and all previous pledges are cancelled

* Undefined: If it is unknown which of the above categories is most appropriate

While the target feature was originally comprised of six classes, the present analysis was only concerned with predicting whether a Kickstarter project will be successful or fail. Furthermore, all of the observations which related to the four other target levels combined only comprised a minority of the overall observations (10.91%). As such, it was decided that observations classifed under the levels of live, canceled, suspended or undefined would not be considered in the analysis, and only observations relating to successful or failed projects would be retained.

```{r}
ks_filt<-ks%>%filter(state %in% c("successful","failed"))
```

As as result of this, the dataset now contained 331,675 observations relating only to successful and failed Kickstarter projects. The target feature would only have two classes and therefore the analysis would be binary classification.

### Descriptive Features

* ID: continuous
* name: string
* category: factor (159 levels)
* main_category: factor (Art, Comics, Crafts, Dance, Design, Fashion, Film & Video, Food, Games, Journalism, Music, Photography, Publishing, Technology, Theater)
* currency: factor (AUD, CAD, CHF, DKK, EUR, GBP, HKD, JPY, MXN, NOK, NZD, SEK, SGD, USD)
* deadline: date
* goal: continuous
* launched: date
* pledged: continuous
* backers: integer
* country: (AT, AU, BE, CA, CH, DE, DK, ES, FR, GB, HK, IE, IT, JP, LU, MX, N,0\", NL, NO, NZ, SE, SG, US)
* us_pledged: continuous
* us_pledged_real: continuous
* us_goal_real: continuous

Many of the descriptive features here are self-explanatory. The last three features are corrections of the pledged and goal features, which convert the relevant currency amounts of the associated observations to USD. Information from the website where the dataset was retrieved suggests that usd_pledged was created to standardise the pledged column to USD, but there were issues with the conversion which resulted in errors. The usd_pledged_real feature amended these errors and has the correct amount pledged in USD. Similarly, the feature usd_goal_real has the USD amount of the project's goal.

Inspection of the levels of the country feature highlight a potential typo, where a country was assigned the code 'N,0\"', which clearly does not correspond with the rest of the levels. This will be examined further in the next section of this report.

## Data Preprocessing

### Initial Inspection and Error Identification

Prior to summarization, some obviously redundant columns were removed. The features ID and name would clearly add no value in predicting the target variable and were dropped. Additionally, as the columns usd_goal_real and usd_pledged_real contained the correct standardised goal and pledged amounts respectively, the pledged, goal and usd_pledged features were no longer relevant and so were dropped.

```{r}
ks_filt$ID<-NULL
ks_filt$name<-NULL
ks_filt$`usd pledged`<-NULL
ks_filt$pledged<-NULL
ks_filt$goal<-NULL
```

The dataset was summarised in Table 1 and 2 below.

```{r}
head(ks_filt)%>% knitr::kable( caption = 'Table 1. First 6 rows of Kickstarter Data')
```


```{r}
summarizeColumns(ks_filt)%>% knitr::kable( caption = 'Table 2. Feature Summary Prior To Data Preprocessing')

```

From Table 1 and Table 2, it is clear that the data was read in correctly and there no missing values or any obvious errors/typos. However, some issues were noted.

* As mentioned previously, there was a suspected typo in the name of one the levels of the country feature - N,0\".

* Looking at the classes of each feature shown in Table 1, the two date columns - launched and deadline - are not of the same class, which may become an issue if these columns are needed to created a new variable. Furthermore, as the time dimension is not applicable as a feature within the present analysis, these two columns will ultimately need to be dropped.

* The catergory and main_category columns appear to represent the same underlying feature - the type of business of the Kickstarter project, however, as seen in Table 2, the cardinality of category was very large at 159 levels compared to that of main_category, which had only 15 levels.

* The currency and country columns seem to hold fairly similar information. As such, dropping one may be necessary. This would need to be investigated visually, however, so will be dealt with in the final section of this report.

* It is clear that all factor columns were defined as the character class when imported and will need be converted appropriately.

### Data Wrangling and Correction

Firstly, the extent of the N,0\" typo within the country feature was explored further. Perusal of the website where data were retreived provided no explaination to the meaning of this value. Therefore, this level was considered to classify as a missing value, and was converted as such. The proportion of complete cases (i.e. those which had no missing values) was then calculated.

```{r}
ks_filt$country <- ifelse(ks_filt$country == "N,0\"", NA, ks_filt$country)
sum(complete.cases(ks_filt))/nrow(ks_filt)
```

Here, we see that the proprotion of complete cases is very high, at 99.94% of all observations. This suggested that the number of observations affected by the typo was minimal and as such, it was deemed appropriate to omit all affected observations from the analysis.

```{r}
ks_filt<-ks_filt%>%na.omit()
```

Next, it was decided that a new feature would be created to represent the duration in days that each project lasted by finding the difference between the two date columns. First, the launched column would have to be converted to the correct date format. In its current state, launched included not only the date, but also the time at which the project was launched, so it was converted.

```{r}
ks_filt$launched <- format(as.POSIXct(ks_filt$launched,format='%Y/%m/%d %H:%M:%S'),format='%Y/%m/%d')
ks_filt$launched<-ks_filt$launched%>%as.Date()
```

The duration feature was then added to the dataset by subtracting the launched column from the deadline column. Subseqently, the launched and deadline columns were dropped as they were no longer needed.

```{r}
ks_filt$duration<-as.numeric(ks_filt$deadline-ks_filt$launched)
ks_filt$launched<-NULL
ks_filt$deadline<-NULL
```

Next, the main_category and category features were considered. Inspection of these features highlighted that main_category was simply be a reduced version of catgeory with far fewer levels, which would be more useful in the upcoming analysis. This left category essentially redundant, so it was removed.

```{r}
ks_filt$category<-NULL
```

Finally, all features loaded in as the character type were converted to factors.

```{r}
ks_filt[sapply(ks_filt, is.character)] <- lapply(ks_filt[sapply(ks_filt, is.character)], as.factor)
```

As a result of all the above amendments, the dataset now contained 331,465 observations, one target feature with two levels and six feature variables.

# Data Exploration

## Univariate Visualisations

Initially, the overall counts of successful and failed projects was examined.

```{r}
p1<-ggplot(ks_filt,aes(x=state))

p1+geom_bar()+
  labs(title="Figure 1. Bar Chart of Project State",
       y="Count",
       x="State")
```

Clearly the main point here is that there were more failed projects in the data than there were sucessful ones. This was considered in visualisations to come.

The main_category feature was then visualised to explore the differences in counts across its levels, as well as the proportions of successful and failed projects within its levels.

```{r}
p2 <- ggplot(ks_filt, aes(x = main_category))

p2 + geom_bar()+
  labs(title = "Figure 2. Bar Chart of Project Category",
       x = "Category",
       y = "Count")+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```


```{r}
cat_state <- table(ks_filt$main_category,ks_filt$state, dnn = c("Category","State"))
prop.cat_state<-data.frame(prop.table(cat_state,1))
colnames(prop.cat_state) <- c("Category","State","Proportion")

p3<-ggplot(prop.cat_state,aes(x=Category,y=Proportion,fill=State))
p3+geom_bar(stat="identity",position="dodge")+
  labs(title="Figure 3. Proportions of Project State within Catgeories",
       y="Proportion within Category")+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

From Figure 2, it can be seen that projects based on film and video as well as music were popular, while dance and journalism based projects were less common. Furthermore, Figure 3 highlights that generally, most categories are associted with failed projects, except for comics, dance, music and theater. Clearly these are all related to the arts.

Next we looked at the currency and country feature simultaniously to identify any redundancy.

```{r}
ggplot(ks_filt, aes(x = country)) + geom_bar() +
  labs(title="Figure 4. Bar Chart of Country",
       y="Count",
       x="Country")
```

```{r}
ggplot(ks_filt, aes(x = currency)) + geom_bar()+
  labs(title="Figure 5. Bar Chart of Currency",
       y="Count",
       x="Country")
```

Inspection of Figures 4 and 5 does highlight a striking similarity. Additionally, they also serve to emphasise that a large proportion of the projects are associated with the US, and also, that many other countries and currencies have much smaller amounts of associated projects. As such, utilising either of these features may lead to some issues with regards to disproportional levels. It was therefore decided that it would be best to aggregate the levels of the country feature to represent the continents, and to then drop the country and currency features. The plots of the new continent feature are shown below.

```{r}
europe <- c("AT", "BE", "DE", "DK", "ES", "FR", "GB", "IE", "IT", "LU", "NL", "NO", "SE")
north_america <- c("CA", "MX", "US")
asia <- c("CH", "HK", "JP", "SG")
oceania <- c("AU", "NZ")

ks_filt$continent <- case_when(
  ks_filt$country %in% europe ~ "Europe",
  ks_filt$country %in% north_america ~ "North America",
  ks_filt$country %in% asia ~ "Asia",
  ks_filt$country %in% oceania ~ "Oceania"
)

ks_filt$country<-NULL
ks_filt$currency<-NULL
```

```{r}
p4 <- ggplot(ks_filt, aes(x = continent))

p4 + geom_bar(position = "dodge")+
  labs(title="Figure 6. Bar Chart of Continent",
       y="Count",
       x="Continent")
```

Obviously, this Figure 5 still shows that North America dominates the projects being assessed, however the other levels are now more concentrated. Despite this, the amount of projects originating from Asia and Oceania remains very small.

```{r}
cont_state <- table(ks_filt$continent,ks_filt$state, dnn = c("Continent","State"))
prop.cont_state<-data.frame(prop.table(cont_state,1))
colnames(prop.cont_state) <- c("Continent","State","Proportion")

p5<-ggplot(prop.cont_state,aes(x=Continent,y=Proportion,fill=State))
p5+geom_bar(stat="identity",position="dodge")+
  labs(title="Figure 6. Proportions of Project State within Continent")
```

From the Figure 6 above, we see that the proportions of failed and succesfull projects from Asia and Europe are almost identical. North America appears to have a slightly higher proportion of successful projects in comparision, while Oceania has a lower proportion of successful projects.

The duration feature was explored next. As seen from Figures 7 and 8 below, there is little distinction between the duration of successful and failed projects. The vast majority of projects have a duration of 30 days and this means that as most projects have the same value. Therefore, the feature essentially becomes useless as there is very little variation. This will need to further investigated in the next phase of this analysis in order to determine whether duration is suitable feature to include.

```{r}
p6 <- ggplot(ks_filt, aes(x = duration))
p6+geom_histogram()+
  labs(title="Figure 7. Histogram of Project Duration",
       x="Duration",
       y="Count")
```


```{r}
p7 <- ggplot(ks_filt, aes(x = duration))

p7 + geom_histogram(aes(fill = state)) + 
labs(title = "Figure 8. Histogram of Project Duration by State",
x = "Duration",
y = "Count",
fill = "State") +
facet_grid(~state)
```

Figure 9 below depicts the number of backers each projects had and shows a clear positive skew. Figure 10 further shows that the distribution is essentially split across the state feature, with failed projects being especially associted with zero backers, and successful ones having much more varience. 
It should be noted that some projects had very large amounts of backers which proved problematic when visualised. As such, these values were filtered out for visualisation purposes, but were retained for analysis as they were valid observations.

```{r}
back_filt<-ks_filt%>%filter(backers<150)

p8<-ggplot(back_filt,aes(x=backers))
p8+geom_histogram()+ 
labs(title = "Figure 9. Histogram of Project Backers",
x = "Backers",
y = "Count")
```


```{r}
p9 <- ggplot(back_filt, aes(x = backers,fill=state))

p9 + geom_density(alpha=.4) + 
labs(title = "Figure 10. Density of Project Backers by State",
x = "Backers",
y = "Density",
fill = "State")
```


Each project's goal was visualised in Figures 11 and 12. Figure 11 highlights an unsual distribution with spikes at every $5000. This may be attributable to indivduals commonly setting goals at these essentially arbitrary points.
Figure 12 highlights an interesting insight which suggests that projects with lower goals are more likely to be successful than projects with higher goals, with the threshold level at approximately $4000. This makes intuitive sense as smaller goals would be more easily achieved, which would then classify the project as successful.
Again, observations with very large goals were fitlered out to best visualise the feature.

```{r}
ks_goal <- ks_filt %>% filter(usd_goal_real <= 25000)

p10 <- ggplot(ks_goal, aes(x = usd_goal_real))

p10 + geom_histogram() + 
labs(title = "Figure 11. Histogram of Project Goal by State",
x = "Goal",
y = "Count")
```


```{r}
p10+ geom_density(aes(fill = state), alpha = 0.4)+ 
labs(title = "Figure 12. Density of Project Goal by State",
x = "Goal (USD)",
y = "Density",
fill = "State")
```

The amount pledged to each project is visualised below in Figures 13 and 14. Here, there is a striking similarity to the visualisations of the backers feature shown in Figures 9 and 10. Again, the overall distribution is highly positively skewed, with the majority of failed projects recieving very little fudning, while the amount pledged to successful projects has considerable variablity. 
The data used in this visualisation was once again filtered of extreme values.

```{r}
pledge_filt<-ks_filt%>%filter(usd_pledged_real<4000)

p11 <- ggplot(pledge_filt, aes(x = usd_pledged_real))

p11 + geom_histogram()+ 
labs(title = "Figure 13. Histogram of Amount Pledged to Project",
x = "Amount Pledged (USD)",
y = "Count")

```

```{r}
p12 <- ggplot(pledge_filt, aes(x = usd_pledged_real,fill=state))

p12 + geom_density(alpha=.4) + 
labs(title = "Figure 14. Density of Amount Pledged to Project by State",
x = "Amount Pledged (USD)",
y = "Density",
fill = "State")
```

## Multivariate Visualisations

Figure 15 below highlights the relationship between a project's goal and its category. Inspection of the different facets emphasises that while setting smaller goals is generally likely to result in a succesful project, this is not true for all categories. Categories such as food, design and technology do not strickly adhere to this rule and demonstrate almost the same amount of succesful projects at both high and low goals. Similarly, other categories such as dance, comics and theatre are also different in that they seem to have high proportions of successful projects regardless of the goal. 


```{r}
p13<-ggplot(ks_goal,aes(x=usd_goal_real, fill=state))

p13 + geom_histogram() + facet_wrap(~main_category)+
  labs(title = "Figure 15. Project Goal by Project State across Category",
       x="Goal (USD)",
       y="Count")
```


```{r}
pledge_filt2<-ks_filt%>%filter(usd_pledged_real<7500)
p14<-ggplot(pledge_filt2,aes(x=continent,y=usd_pledged_real,fill=state))
p14+geom_boxplot()+
  labs(title = "Figure 16. Boxplot of Amount Pledged by Project State and Continent",
       x="Continent",
       y="Amount Pledged (USD)")
```

The amount pledged to a project according to its continent of origin is depicted in Figure 16 above. Here we see minimal differences in the distribution of the amounts pledged to failed projects across continents, all showing on average next to zero funding. It is also clear that there are many outliers in the failed state, which despite receiving considerable funding still failed. However, it is interesting to see that Asia has the highest mean amount pledged to successful projects of all other continents, when it was previously shown that it had by far the smallest representation in the dataset. This suggests that in general, projects started in Asia would receieve more funding.

```{r}
ks_scatter<-ks_filt%>%filter(backers < 10000)
ks_scatter<-ks_scatter%>%filter(usd_pledged_real < 1000000)

set.seed(90)
index <- sample(1:nrow(ks_scatter), 5000)
ks_rand<-ks_scatter[index, ]

p15 <- ggplot(ks_rand, aes(x = backers, y = usd_pledged_real))

p15 + geom_point(aes(colour = state,size=usd_goal_real),alpha=.3)+
  facet_wrap(~state)+
  labs(title="Figure 17. Project State by Amount Pledged to Project, Backers and Goal",
       y="Amount Pledged (USD)",
       x="Backers")

```

There are several interesting points to note about Figure 17. Unsurprisingly, failed projects can be seen to be highly concertrated around the origin, suggesting they received little funding and backers. Visually, this gives off the sense that the failed facet has fewer points, when in fact failed projects were more common than succesful ones. The successful projects are much more spread out and suggest a moderate, postiive relationship between backers and amount pledged. The dispertion of successful points is also promising for the upcoming analysis, as it suggests that there is not too much correspondence between backers and amount pledged. However, the concentration of points does indicate that succesful projects typically receive few backers and and moderate funding. This may ultimately be linked to their smaller goals which contrast the much larger goals evident among failed projects. As previously mentioned, this would suggest that perhaps failed projects were too ambitious with their goals, and considering this, it would make sense that the projects with smaller goals were more likely to reach their targets and become successful; requiring fewer backers pledging smaller amounts. 
This plot was constructed using a random sample from the dataset with extreme values filtered from the backers and usd_pledged_real features.

# Summary

In this phase of the report, the relevance of each feature in the original Kickstarter dataset was assessed in order to best predict whether a project would fail or succeed. Of the 15 original features, only five were retained, and two new features were created. For the target feature, only observations relating to successful and failed projects were included and others were dropped. ID and name were both descriptive features which would add no insight into the upcoming analysis so were dropped. The pledged, goal and usd_pledged features were all dropped as they contained ultimately useless information. Deadline and launched were combined to produce the duration feature, and then were both dropped. The category feature was also dropped as it was found to be redunadant. Finally, some typos were found within the country feature and all relevant observations were dropped. 

Visual assessment of the remaing features further demonstrated the usefulness of some features. Namely, it was shown that the cardinality of the country feature was too high given the low counts of projects within some coutnries and as such, the country feature was dropped and converted to represent continent. Additionaly the duration feature was found to add little insight into the distinction between failed and successful projects, but it was retained for now. Important insights drawn from visualising remaining features included how projects with smaller, realistic goals are more likely to succeed, however this is not strictly true for all categories - some of which, such as those related to arts, are highly successful regardless of goal. It was found that North American projects were the most common and also most likely to be successful and that Asian projects, which were least common, received the most funding. Finally, it was found that although successful projects are unsprisingly more likely to have more backers and funding than failed ones, the relationship between these features suggests most succesful projects receive relatively few backers who pledge moderate amounts, likely in accordance with their smaller goals.